---
title: "Creating a TI method using docker"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating a TI method using docker}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, echo=FALSE, message=FALSE}
library(dplyr)
library(readr)
library(purrr)
```

This vignette describes how to include your own method into the dynverse workflow. We'll illustrate this with a very basic method: using one of the components of a PCA as the pseudotime.

There are two main ways to add a new method:

* **Directly within R**. In this case you create an R function or R packages. The user will have to install all dependencies. This is use case is described in another vignette.
* **Using a docker**. In this case you create a docker which uses some input files (eg. a file containing the expression) and returns some output files. This use case is described here.

A TI docker requires at least three files. The files for this example are also available at [https://github.com/dynverse/comp1](https://github.com/dynverse/comp1).

* A definition yaml or json file, used to determine the format of input, output and parameters.
* A Dockerfile, used to construct the docker.
* Script(s) to load the data, run the method, and save the output

## Method definition file (`definition.yml`)

An example of a `definition.yml` file (as can be found on the [github](https://github.com/dynverse/comp1/blob/devel/R_feather/definition.yml)):

```{r, echo=FALSE, results="asis"}
read_file("https://raw.githubusercontent.com/dynverse/comp1/devel/R_feather/definition.yml") %>%
  paste0("```yaml\n", ., "\n```") %>% 
  cat()
```

This yaml file contains:

#### name
A name of the method

#### parameters

```{r, echo=FALSE}
parameter_types <- tribble(
  ~type, ~properties, ~example,
  "numeric", "lower, upper, distribution, mean, sd, rate", "3.5",
  "numeric_vector", "lower, upper, distribution, mean, sd, rate, length", "[3.5, 4.5]",
  "integer", "lower, upper, distribution, mean, sd, rate", "3",
  "integer_vector", "lower, upper, distribution, mean, sd, rate, length", "[3, 10]",
  "logical", "", "true",
  "logical_vector", "length", "[true, false, true]",
  "discrete", "values", "mds",
  "discrete_vector", "length, values", "[mds, pca, tsne]"
)
```

A named list of parameters for the method. A parameter always has a type and a default value.  Optionally, parameters can also have other properties which define the parameter space. These are used for parameter tuning:

```{r, echo=FALSE}
parameter_types %>% knitr::kable()
```

- **distribution** The distribution from which to sample numeric or integer parameters. Can be:
    - `uniform` (default)
    - `normal` (also specify **mean** and **sd**)
    - `exponential` (also specify **rate**). 
- **lower** and **upper** The lower and upper bounds. Can be used with any distribution.
- **values** A list of possible values for discrete parameters
- **length** Length of the parameter for vector parameters

#### input


The input determines which kind of expression and prior information is given to the method. It contains the following properties:

- **format** The format of the input, which can be:
```{r, echo=FALSE}
comp1_url <- function(name, x) glue::glue("[{name}](https://github.com/dynverse/comp1/tree/devel/{x})")
comp1_urls <- function(names, xs) glue::collapse(map2_chr(names, xs, comp1_url), ", ")
input_url <- function(name, x) glue::glue("[{name}](https://github.com/dynverse/comp1/tree/devel/inst/example_inputs/{x})")
inputs_url <- function(names, xs) glue::collapse(map2_chr(names, xs, input_url), ", ")

tribble(
  ~output_format, ~description, ~example_script,
  "hdf5", "HDF5 format", comp1_urls(c("python", "R"), c("python_hdf5", "R_hdf5")), 
  "text", "csv and json files", comp1_urls(c("python", "R"), c("python_text", "R_text")),
  "feather", "[Feather](https://github.com/wesm/feather) dataframe files", comp1_urls(c("python", "R"), c("python_feather", "R_feather")),
  "rds", "R binary files", comp1_urls(c("R"), c("R_rds"))
) %>% knitr::kable()
```


- **required** inputs (eg. expression) and/or **optional** inputs (eg. start_id):

```{r, echo=FALSE}
data(allowed_inputs, package="dynwrap")
allowed_inputs <- allowed_inputs %>% filter(input_id != "dataset")

# find example files
found_input_example_files <- list.files(devtools:::shim_system.file("example_inputs/text/", package="dynwrap"))
input_examples <- set_names(
  found_input_example_files, 
  tools::file_path_sans_ext(found_input_example_files)
)

# wrap example files around a url
input_example_url <- function(input_id) {
  if(input_id %in% names(input_examples)) {
    glue::glue("[{input_id}](https://github.com/dynverse/dynwrap/blob/master/inst/example_inputs/text/{input_examples[[input_id]]})")
  } else {
    input_id
  }
}

# put everything in a table
allowed_inputs %>% 
  mutate(
    input_id = map_chr(input_id, input_example_url)
  ) %>% 
  select(
    input_id, 
    description
  ) %>% 
  knitr::kable()

rm(allowed_inputs)
```

```{r, echo=FALSE}
github_url <- function(x, prefix) {
  glue::glue("[{x}](https://github.com/dynverse/dynwrap/blob/master/{prefix}{x})")
}
```


#### output

The output determines how the output of a method can be transformed into a common trajectory model.

- **format** The format of the output, which can be

```{r, echo=FALSE}
comp1_url <- function(name, x) glue::glue("[{name}](https://github.com/dynverse/comp1/tree/devel/{x})")
comp1_urls <- function(names, xs) glue::collapse(map2_chr(names, xs, comp1_url), ", ")

tribble(
  ~output_format, ~description, ~example_script,
  "text", "csv and json files", comp1_urls(c("python", "R"), c("python_text", "R_text")),
  "feather", "[Feather](https://github.com/wesm/feather) dataframe files", comp1_urls(c("python", "R"), c("python_feather", "R_feather")),
  "rds", "R binary files", comp1_urls(c("R"), c("R_rds")),
  "dynwrap", "directly wrapped using [dynwrap](https://github.com/dynverse/dynwrap) functions", comp1_urls(c("R"), c("R_dynwrap"))
) %>% knitr::kable()
```

- **outputs** The kinds of output this wrapper produces. This always include at least one output which contains enough information to produce a trajectory:

```{r, echo=FALSE}
data(allowed_outputs, package = "dynwrap")

# find example files
found_output_example_files <- list.files(devtools:::shim_system.file("example_outputs/text/", package="dynwrap"))
output_examples <- set_names(
  found_output_example_files, 
  tools::file_path_sans_ext(found_output_example_files)
)

# wrap example files around a url
output_example_url <- function(output_id) {
  if(output_id %in% names(output_examples)) {
    glue::glue("[{output_id}](https://github.com/dynverse/dynwrap/blob/master/inst/example_outputs/text/{output_examples[[output_id]]})")
  } else {
    output_id
  }
}
output_example_urls <- function(output_ids) {
  if (length(output_ids) > 0) {
    map(output_ids, output_example_url) %>% 
      glue::collapse("<br>") %>% 
      stringr::str_replace("(.*milestone_percentages.*)<br>(.*progressions.*)", "\\1 or \\2") # put or in between percentages and progressions
  } else {
    ""
  }
}

# put everything in a table
outputs <- allowed_outputs %>% 
  mutate(
    required = map_chr(required_args, output_example_urls), 
    optional = map_chr(optional_args, output_example_urls)
  ) %>% 
  select(
    output_id, 
    description,
    required,
    optional,
    creates_trajectory
  )
```

```{r, echo=FALSE}
outputs %>%
  filter(creates_trajectory) %>% 
  select(-creates_trajectory) %>% 
  knitr::kable()
```


In addition, a method can also generate other output which will be added to the model. This includes a grouping/clustering (`grouping`) or a dimensionality reduction (`dimred`).

```{r, echo=FALSE}
outputs %>%
  filter(!creates_trajectory) %>% 
  select(-creates_trajectory) %>% 
  knitr::kable()
```

#### Other information about the method

The `definition.yml` file can also contain other information about the methods, such as the `authors`, the detectable `trajectory_types`, the url of the code, the `version`, ... See the [dynmethods containers](https://github.com/dynverse/dynmethods/tree/master/containers) for examples.

## The `Dockerfile`

An example of a `Dockerfile` (as can be found on the [github](https://github.com/dynverse/comp1/blob/devel/R_feather/Dockerfile)):

```{r, echo=FALSE, results="asis"}
read_file("https://raw.githubusercontent.com/dynverse/comp1/devel/R_feather/Dockerfile") %>%
  paste0("```Dockerfile\n", ., "\n```") %>% 
  cat()
```

In this example, we add all the files present in the current working directory (which contains the run.R file) to the /code directory.

All dependencies of the method should be installed in this Dockerfile (or from parent docker images)

This file should always contain an entrypoint, which will be called when the TI method is run.

## Doing trajectory inference

All input files requested in the `definition.yml` will be mounted in the `/ti/input` directory of the docker. After inferring a trajectory, the method should write output files into the `/ti/output` directory. An example (as can be found on the [github](https://github.com/dynverse/comp1/blob/devel/R_feather/run.R)):

```{r, echo=FALSE, results="asis"}
read_file("https://raw.githubusercontent.com/dynverse/comp1/devel/R_feather/run.R") %>%
  paste0("```r\n", ., "\n```") %>% 
  cat()
```

## Testing the TI method

Now we're all set to test the TI method on data. 

We first have to build the method to an image. Make sure that docker is installed (https://docs.docker.com/install/). Next, build the container in shell

```bash
docker build -t dynverse/comp1 .
```

or in R

```r
system("docker build -t dynverse/comp1 .")
```

The `-t dynverse/comp1` argument gives a tag to this container, so that it can be easily run later.

Although this container can be run anywhere as long as the `/ti/input` and `/ti/output` folders are mounted, we'll run it here using the `dynwrap` workflow. First, we have to wrap the docker into a method function.
```r
library(dynwrap)
ti_comp1 <- create_ti_method_with_container("dynverse/comp1")
```

```{r, echo=FALSE}
library(dynwrap)
```

If the method is on the docker hub, it can also be pulled
```r
ti_comp1 <- create_ti_method_with_container("dynverse/comp1")
```

```{r, echo=FALSE}
source(devtools:::shim_system.file("tests/testthat/helper-ti_comp1.R", package = "dynwrap"))
```

Then we'll use some test data
```{r}
data(example_dataset)
dataset <- example_dataset
```

Now infer a trajectory
```{r}
model <- infer_trajectory(dataset, ti_comp1())
```

[dynplot](https://github.com/dynverse/dynplot) can be used to easily visualise the trajectory model:
```r
# devtools::install_github("dynverse/dynplot")
dynplot::plot_dimred(model, color_cells = "pseudotime" , expression_source = dataset$expression)
```

```{r echo=FALSE}
if ("dynplot" %in% rownames(installed.packages())) {
  dynplot::plot_dimred(model, color_cells = "pseudotime" , expression_source = dataset$expression)
}
```

We can also provide some optional prior information to the method:
```{r}
model <- infer_trajectory(dataset, ti_comp1(), give_priors = "start_id")
```

```r
dynplot::plot_dimred(model, color_cells = "pseudotime" , expression_source = dataset$expression)
```

```{r echo=FALSE}
if ("dynplot" %in% rownames(installed.packages())) {
  dynplot::plot_dimred(model, color_cells = "pseudotime" , expression_source = dataset$expression)
}
```

Or change some parameters:

```{r}
model <- infer_trajectory(dataset, ti_comp1(component = 2))
```

```r
dynplot::plot_dimred(model, color_cells = "pseudotime" , expression_source = dataset$expression)
```

```{r echo=FALSE}
if ("dynplot" %in% rownames(installed.packages())) {
  dynplot::plot_dimred(model, color_cells = "pseudotime" , expression_source = dataset$expression)
}
```

## Debugging the container

The container can be debugged by running the method with `debug = TRUE`:

```r
infer_trajectory(dataset, ti_comp1, debug = TRUE)
```

This will print a shell command which can be used to enter the container with the /ti/input and /ti/output directories mounted.

## Making your method available for other users

The easiest way to make your method available to other users is by [pushing the docker to docker hub](https://docs.docker.com/docker-cloud/builds/push-images/):

```bash
docker push dynverse/comp1
```

Users will then be able to run the method by specifying the docker id:

```{r, eval=FALSE}
model <- infer_trajectory(dataset, "dynverse/comp1")
```

If you wish to automatically update the underlying docker to a certain version, you can specify the digest as follows:
```{r, eval=FALSE}
model <- infer_trajectory(dataset, "dynverse/comp1@sha256:f1e406c5a977a66eb5bdcdced615ddf39f0d05b308aa01092f69d81db0eb8a0a")
```

We encourage developers to submit their TI method to the 
[dynmethods R package](https://github.com/dynverse/dynmethods),
which presents TI methods wrapped in docker containers to users in a user-friendly format. 
Feel free to [create an issue](https://github.com/dynverse/dynmethods/labels/new%20method)
or [send us a pull request](https://github.com/dynverse/dynmethods/pulls).
