---
title: "Creating a TI method: Script"
date: "`r Sys.Date()`"
author:
- Wouter Saelens
- Robrecht Cannoodt
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating a TI method: Script}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
dir <- tempdir()
setwd(dir)
knitr::opts_knit$set(root.dir = normalizePath(tempdir(), winslash = '/'))
```


```{r setup}
library(dynwrap)
```

An alternative to wrapping a script inside R, is to wrap it using an external script. Because this does not provide any dependency management, this is not really useful for method end-users, but rather as a way to easily develop a TI method and to ultimately step up towards [containerised wrapping](../create_ti_method_container).

Similarly as a wrapper written in R, you'll need to provide both a definition (= a `definition.yml`) and a way to run the methods (= a script).

## `definition.yml`

The definition has the same hierarchical structure as used by the `definition()` function. A minimal example 

```{r, echo = FALSE}
definition_string <- "method:
  id: comp_1

parameters:
  - id: component
    default: 1
    type: integer
    distribution:
      type: uniform
      lower: 1
      upper: 10
    description: The nth component to use

wrapper:
  input_required: expression
  input_optional: start_id
"

readr::write_file(definition_string, "definition.yml")
knitr::asis_output(paste0("```yaml\n", definition_string, "\n```"))
```

Check out the wrappers in dynmethods for an overview of what can be included in the definition file. For example: https://github.com/dynverse/ti_paga/blob/master/definition.yml

## Running script

A running script reads in the data, creates a trajectory, and again writes the trajectory to a file. To do the loading and writing tasks, we provide helper libraries for R and python: [dyncli](https://github.com/dynverse/dyncli) and [dynclipy](https://github.com/dynverse/dynclipy).

The wrapper script will typically have the following structure:

- A shebang, such as `#!/usr/bin/env Rscript` or `#!/usr/bin/env python`
- A call to dyncli(py) to load in the data, such as `dataset <- dyncli::main()` or `dataset = dynclipy.main()`
- All you need to infer the trajectory
- Constructing the trajectory model using [any of the wrapping methods](../create_ti_method_wrappers)
- A call to dyncli(py) to write the data, such as `dyncli::write_output(trajectory)` or `trajectory.write_output()`

A minimal example script for R:

```{r, echo = FALSE}
run_r_string <- "#!/usr/bin/env Rscript

dataset <- dyncli::main()

library(dynwrap)
library(dplyr)

# infer trajectory
pca <- prcomp(dataset$expression)

pseudotime <- pca$x[, dataset$parameters$component]

# flip pseudotimes using start_id
if (!is.null(dataset$priors$start_id)) {
  if(mean(pseudotime[start_id]) > 0.5) {
    pseudotime <- 1-pseudotime
  }
}

# build trajectory
trajectory <- wrap_data(cell_ids = rownames(dataset$expression)) %>%
  add_linear_trajectory(pseudotime = pseudotime)
  
# save output
dyncli::write_output(trajectory, dataset$output)
"

readr::write_file(run_r_string, "run.R")
knitr::asis_output(paste0("```r\n", run_r_string, "\n```"))
```

Make sure this script is executable!

```{bash}
chmod +x run.R
```


and for Python:

```{r, echo = FALSE}
run_py_script <- "#!/usr/bin/env python

import dynclipy
dataset <- dynclipy.main()

import sklearn.decomposition

# infer trajectory
pca = sklearn.decomposition.PCA()
dimred = pca.fit_transform(dataset['expression'])
pseudotime = dimred[:, dataset['parameters']['component']]

# build trajectory
trajectory <- dynclipy.wrap_data(cell_ids = dataset['expression'].index)
trajectory.add_linear_trajectory(pseudotime = pseudotime)

# save output
trajectory.write_output(dataset['output'])
"

readr::write_file(run_py_script, "run.py")
knitr::asis_output(paste0("```python\n", run_py_script, "\n```"))
```

## Testing it out

```{r}
method <- create_ti_method_definition(
  "definition.yml",
  "run.R"
)
dataset <- dynwrap::example_dataset
trajectory <- infer_trajectory(dataset, method(), verbose = TRUE)
```

```{r}
if ("dynplot" %in% rownames(installed.packages())) {
  dynplot::plot_dimred(trajectory, color_cells = "pseudotime" , expression_source = as.matrix(dataset$expression))
}
```

## Interactive development

With the debug parameter, it is possible to enter the script interactively. This makes it easier to develop TI method as you can just load in the input data, and code!

```{r, error = TRUE}
trajectory <- infer_trajectory(dataset, method(), debug = TRUE)
```

## Making your method available for other users

Wrapping a method inside a script does not have any dependency management, and is therefore only meant for development purposes. To deploy your method to other users, check out the [containerisation tutorial](../create_ti_method_container)!

## Wrapping a method without dyncli(py)

While using dyncli to load and save the data is the most straightforward when using R and python, it's also possible to 
