---
title: "Creating a TI method: Types of wrappers"
date: "`r Sys.Date()`"
author:
- Wouter Saelens
- Robrecht Cannoodt
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating a TI method: Types of wrappers}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, echo = FALSE, message = FALSE}
library(tibble)
library(dynwrap)
```

dynwrap always represents trajectories in the same way, as illustrated here with a tree trajectory

- Milestone network, contains information of connections between milestones

```{r, echo = FALSE}
tribble(
  ~from, ~to, ~length, ~directed,
  "A", "B", 1, FALSE,
  "B", "C", 2, FALSE,
  "B", "D", 1, FALSE,
  "C", "E", 1, FALSE,
  "C", "F", 1.5, FALSE
)
```

- Milestone percentages, contains the positions of cells between milestones. For each cell, the percentages sum to one
```{r}
cell_ids <- letters
# progressions <- 
```

- Progressions, an alternative to milestone percentages, also contains the positions of each cell but now based on its progression through an edge
```{r}
head(dynwrap::example_dataset$progressions)
```

- Divergence regions, contain the information on milestones

```{r}
head(dynwrap::example_dataset$divergence_regions)
```
